---
title: "EDA_Analysis"
output: html_document
date: "2025-07-17"
---
### I. INTRODUCTION

### II. Dataset Manipulation

#### 1) Selecting Response Variables and Predictors
```{r}
# Install packages
install.packages("ggplot2")
install.packages("car")
install.packages("carData")
install.packages("ResourceSelection")
install.packages("pROC")
install.packages("dplyr")
```

```{r}
# Load libraries
library(ggplot2)
library(car)
library(carData)
library(ResourceSelection)
library(pROC)
library(dplyr)
```

```{r}
#Load original dataset
df <- read.csv("data/2015.csv")
```

```{r}
#Select relevant predictors and dependent variable
df1 <- df[, c("CVDINFR4", "SEX", "INCOME2", "X_RACEGR3", "X_AGEG5YR", "X_BMI5",
              "X_BMI5CAT", "BPHIGH4", "HLTHPLN1", "EXERANY2")]
```

#### 2) Recoding Variables
Recoding numerical values to detailed categorical levels
```{r}
# Recode CVDINFR4
df1$CVDINFR4[df1$CVDINFR4 %in% c(7, 9)] <- NA
df1$CVDINFR4 <- factor(df1$CVDINFR4, levels = c(1, 2), labels = c("Yes", "No"))

# Recode SEX
df1$SEX <- factor(df1$SEX, levels = c(1, 2), labels = c("Male", "Female"))

# Recode INCOME2
df1$INCOME2[df1$INCOME2 %in% c(77, 99)] <- NA
df1$INCOME2 <- factor(df1$INCOME2, levels = c(1:8),
                      labels = c("<$10k", "$10-15k", "$15-20k", "$20-25k", "$25-35k",
                                 "$35-50k", "$50-75k", ">=75k"))

# Recode _RACEGR3 as factor with appropriate labels
df1$RACEGR3 <- factor(df1$`X_RACEGR3`,
                         levels = c(1, 2, 3, 4, 5, 6, 7, 9),
                         labels = c("White", "Black", "Asian",
                                    "NH/PI", "AI/AN", "Other",
                                    "Multiracial", NA))

# Recode _AGEG5YR
df1$AGEG5YR <- factor(df1$`X_AGEG5YR`, levels = 1:13,
                         labels = c("18–24", "25–29", "30–34", "35–39", "40–44", "45–49",
                                    "50–54", "55–59", "60–64", "65–69", "70–74", "75–79", "80+"))

# Recode _BMI5 
df1$`X_BMI5`[df1$`X_BMI5` == 9999] <- NA

# Recode _BMI5CAT
df1$BMI5CAT <- factor(df1$`X_BMI5CAT`, levels = 1:4,
                         labels = c("Underweight", "Normal", "Overweight", "Obese"))

# Recode BPHIGH4
df1$BPHIGH4[df1$BPHIGH4 %in% c(7, 9)] <- NA
df1$BPHIGH4 <- factor(df1$BPHIGH4, levels = c(1, 2, 3, 4),
                      labels = c("Yes", "No", "Borderline", "Pregnancy"))

# Recode HLTHPLN1
df1$HLTHPLN1[df1$HLTHPLN1 %in% c(7, 9)] <- NA
df1$HLTHPLN1 <- factor(df1$HLTHPLN1, levels = c(1, 2), labels = c("Yes", "No"))

# Recode EXERANY2
df1$EXERANY2[df1$EXERANY2 %in% c(7, 9)] <- NA
df1$EXERANY2 <- factor(df1$EXERANY2, levels = c(1, 2), labels = c("Yes", "No"))
```

#### 3) Handling Missing Values
Filtering out rows with N/A
```{r}
# Remove rows with any NA values
df_clean <- na.omit(df1)  
nrow(df_clean)

#save cleaned files into csv
write.csv(df_clean, "data/cleaned_data.csv", row.names = FALSE)
```

### III. Data Visualization and Tendency Measurement

#### 1) Frequency Distribution of Heart Attack and Predictors

Define Mode Functions for central tendency measurement
```{r}
# Function to calculate mode
get_mode <- function(v) {
  tbl <- table(v)
  mode_val <- names(tbl)[which.max(tbl)]
  return(mode_val)
}
```

**(a)** Response Variable -- CVDINFR4
```{r}
#table
table(df_clean$CVDINFR4)
prop.table(table(df_clean$CVDINFR4))

#Bar graph
ggplot(df_clean, aes(x = CVDINFR4)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of CVDINFR4 Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Head_Attack\n")
cat("Mode:", get_mode(df_clean$CVDINFR4), "\n\n")
```

**Conclusion**: Based on the mode calcularion, No is dominantly observed over Yes, meaning more people who filled out the questionnaire doesn’t have heart attack. However, there is still a sufficient amount of observations having diabetes, with an occurrence of 18,674, which is about 5.85% of the total sample size. 
<br><br>

**(b)** Predictor -- Sex
```{r}
#table
table(df_clean$SEX)
prop.table(table(df_clean$SEX))

#Bar graph
ggplot(df_clean, aes(x = SEX)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Sex Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Sex\n")
cat("Mode:", get_mode(df_clean$SEX), "\n\n")
```

**(c)** Predictor -- Income
```{r}
#table
table(df_clean$INCOME2)
prop.table(table(df_clean$INCOME2))

#Bar graph
ggplot(df_clean, aes(x = INCOME2)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Income Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode and Median
cat("Income\n")
cat("Mode:", get_mode(df_clean$INCOME2), "\n")

# Median (convert to numeric index and retrieve label)
income_numeric <- as.numeric(df_clean$INCOME2)
income_levels <- levels(df_clean$INCOME2)
income_median_label <- income_levels[median(income_numeric, na.rm = TRUE)]
cat("Median:", income_median_label, "\n")
```

**(d)** Predictor -- Race
```{r}
#table
table(df_clean$RACEGR3)
prop.table(table(df_clean$RACEGR3))

#Bar graph
ggplot(df_clean, aes(x = RACEGR3)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Race Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Race\n")
cat("Mode:", get_mode(df_clean$RACEGR3), "\n\n")
```

**(e)** Predictor -- Age
```{r}
#table
table(df_clean$AGEG5YR)
prop.table(table(df_clean$AGEG5YR))

#Bar graph
ggplot(df_clean, aes(x = AGEG5YR)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Age Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode and Median
cat("AGEG5YR\n")
cat("Mode:", get_mode(df_clean$AGEG5YR), "\n")
median_index <- median(as.numeric(df_clean$AGEG5YR), na.rm = TRUE)
median_label <- levels(df_clean$AGEG5YR)[median_index]
cat("Median:", median_label, "\n\n")
```

**(f)** Predictor -- BMI
```{r}
#Bar graph
ggplot(df_clean, aes(x = X_BMI5)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of BMI Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode and Median and Mean
cat("_BMI5\n")
cat("Mode:", get_mode(df_clean$`X_BMI5`), "\n")
cat("Median:", median(df_clean$`X_BMI5`, na.rm = TRUE), "\n")
cat("Mean:", mean(df_clean$`X_BMI5`, na.rm = TRUE), "\n\n")
```

**(g)** Predictor -- Blood Pressure
```{r}
#table
table(df_clean$BPHIGH4)
prop.table(table(df_clean$BPHIGH4))

#Bar graph
ggplot(df_clean, aes(x = BPHIGH4)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Blood Pressure Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Blood_Pressure\n")
cat("Mode:", get_mode(df_clean$BPHIGH4), "\n\n")
```

**(g)** Predictor -- Healthcare Coverage
```{r}
#table
table(df_clean$HLTHPLN1)
prop.table(table(df_clean$HLTHPLN1))

#Bar graph
ggplot(df_clean, aes(x = HLTHPLN1)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Healthcare Coverage Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Healthcare_Coverage\n")
cat("Mode:", get_mode(df_clean$HLTHPLN1), "\n\n")
```

**(h)** Predictor -- EXERANY2
```{r}
#table
table(df_clean$EXERANY2)
prop.table(table(df_clean$EXERANY2))

#Bar graph
ggplot(df_clean, aes(x = EXERANY2)) +
  geom_bar(fill = "#1a80bb", color = "black") +
  ggtitle("Bar Plot of Exercise Distribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Central Tendency -- Mode
cat("Exercise\n")
cat("Mode:", get_mode(df_clean$EXERANY2), "\n\n")
```



